Synthesis of Linear Functional Programs

### 2021-05-08 article/article.tex
uns tweaks

\mypara{Goals}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

@@

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures \dots. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting the
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the \dots \todo{Identity function!}

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus, and their proofs are
exactly the \dots.
The main observations made are ``propositions as types'', ``proofs as
programs'', ``proof checking as type checking''~\cite{}.

Inference in ND are categorized as introduction or elimination, these
correspond to constructors and destructors in programming languages
(e.g. function abstraction vs application, construction of a pair vs
projection).

We introduce a few typing rules to display how a type system can be
formalized.  In natural deduction, there are both \emph{introduction}
and \emph{elimination} rules for almost all logical
connectives.\todo{se mencionares intro e elim antes nao precisas de
  repetir aqui.}  The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
  \]
  When seen as a typing rule for the $\lambda$-calculus, it
  corresponds to the rule for typing variables -- variable $u$ has
  type $A$ if the typing environment contains a variable $u$ of
  type $A$.

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
 In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
 So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
 to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
it useable more than once. To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.

>>>>

\mypara{Goals}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize a term of type $T$?) and time
measurements (i.e. how fast can we synthesize a term of type $T$?').

@@

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures \dots. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting the
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the term $\lambda x{:}A.x$ denotes the
identity function.

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus, and their proofs are
exactly the programs with the given type.
The main observations made are ``propositions as types'', ``proofs as
programs'', ``proof checking as type checking''~\cite{}.

Inference rules in natural deduction are generally categorized as
introduction or elimination rules, which correspond, respectively, to
constructors and destructors in programming languages (e.g.~function
abstraction vs application, construction of a pair vs projections).

We introduce a few rules to both show how a type system can be
formalized and to show the relationship with (intuitionistic)
propositional logic.  In natural deduction, there are both
\emph{introduction} and \emph{elimination} rules for almost all
logical connectives.\todo{se mencionares intro e elim antes nao
  precisas de repetir aqui.}  The following rule captures the nature
of a hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
  \]
  When seen as a typing rule for the $\lambda$-calculus, it
  corresponds to the rule for typing variables -- variable $u$ has
  type $A$ if the typing environment contains a variable $u$ of
  type $A$.

@@

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since
it'll alter the judgment form, and it's the one that ensures logical
strength -- i.e. we're able to translate intuitionistic logic into
linear logic.  The proposition $\bang A$ (read \emph{of course} $A$)
is used (under certain conditions) to make a resource ``infinite''
i.e. to make it useable an arbitrary number of times. To distinguish the
``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
that can only be used once.
